Sequential rank agreement methods for comparison of lists

### 2015-08-10 latex/paper.Rnw
final application revision

Sequential rank agreement can be generalized to incomplete/censored
lists in the following way. Let $\Lambda_P$ be a set of $P$ items and
$\Lambda_l\subset \Lambda_P$ the subset of $d_l$ items that occur in
list $l$.  The case where all lists are censored at depth $d$
corresponds to $d_1=\cdots=d_L=d$. For incomplete/censored lists the
rank function becomes
\begin{equation}
  \tilde R_l(X_p) = \left\{\begin{array}{cl} \{R_l(X_p)\} & \text{ for } X_p\in \Lambda_l \\
      \{d_l+1,\dots,P\} & \text{ for } X_p \not\in \Lambda_l\end{array}\right.
\end{equation}
where we only know that the rank for the unobserved items in list $l$
must be larger than the largest rank observed in that list.

@@

Alternatively, we can remove the restriction on the independence among
the methods by only requiring that there is no information in the
ranking but that the rankings are all based on applying the
method/approaches to the same data
\begin{eqnarray*}
\tilde H_0 & :&  \text{The list rankings are based on data containing}\\
& &   \text{no association to the outcome.}
\end{eqnarray*}
$H_0$ is a quite unrealistic null hypothesis but we can easily obtain
realizations from that null hypothesis simply by permuting the items
within each list and then computing the sequential rank agreement for
the permuted lists. In the uncensored case each experiment contains
$L$ lists of random permutations of the numbers $1,\dots,P$. For the
censored/incomplete case we first permute the numbers $1,\dots,P$ and
then censor list $l$ at $d_l$. The sequential rank agreement curve
from the the original lists can then be compared to the pointwise rank
agreements obtained under $H_0$.

To obtain the distribution under $\tilde H_0$ the idea is to repeat
many times the ranking procedures in uncorrelated data.  Thus, we
first permute the outcome variable of the data set on which the
rankings are based. This removes any association between the predictor
variables and the outcome. Then we apply the methods to the permuted
data set to generate $L$ new rankings and subsequently compute the
sequential rank agreement. Note that we only permute the outcomes and
preserve the structure of the candidate predictors. The randomization
approach requires that we have the original data available and as such
it may not be possible to evaluate $\tilde H_0$ in all situations.

If the sequential rank agreement for the original data lies
substantially below the distribution of the sequential rank agreements
obtained under either $H_0$ or $\tilde H_0$ then this suggests that
the original ranked lists agree \emph{more} than what we would expect
in data with no information.

For the analysis of the Golub data described in Section
\ref{sec:amfol}, Figure~\ref{fig:example1} shows the empirical
distributions of sequential rank agreement under $H_0$ and
$\tilde H_0$ each based on $400$ permutations. Not surprisingly, the
sequential rank agreement under $\tilde H_0$ is lower than the
sequential rank agreement under $H_0$ because the four methods used to
rank the data ($t$ test, logistic regression, elastic net, and MIC)
generally tend to identify (and rank high) similar predictors even if
there are only spurious associations. The two bottom panels in
Figure~\ref{fig:example1} also indicate that the observed sequential
rank agreement (the black line) is better than what would be expected
by chance for data that contain no information since it lies below the
reference areas. However, the censored data also suggests that there
may be at most 1 or 2 ranked items towards the top of the lists that
yield a result better than what would be expected (the bottom-right
plot).

It is important to stress that neither $H_0$ nor $\tilde H_0$ are
related to questions regarding the association between the outcome and
the predictors in the data set. Both hypotheses are purely considering
how the rankings agree in a situation where there is no relevant
information available in the data used for creating the rankings.

@@

A not expected shape of the agreement curves is seen for the Ridge
models for all three tuning criteria. They all show higher
disagreement in the top of the lists followed by a decrease. The
reason behind this behavior is rather subtle. Looking at the
distribution of the regression coefficients we see that a large
proportion of them are numerically very close to zero and have almost
equal absolute value. This is a general feature of the Ridge models in
this data set and seen for all the 1000 trained models.
This implies that when predictors are ranked according to the
magnitude of their coefficients, their actual order becomes more
uncertain and more close to a random permutation. This problem can be
alleviated by artificially censoring all predictors with absolute
coefficient values below a given threshold. For the Ridge models tuned
with the deviance criterion, Figure \ref{fig:app2} (left) shows the
the sequential rank agreement where for each of the 1000 trained
models predictors were artificially censored when their absolute
coefficient value was lower than the $0.1\%$ quantile of the 5000
absolute coefficient values. The corresponding curve from Figure
\ref{fig:app1} (left panel) is shown for comparison.
\textbf{CHECK UP: CENSOR ON POOLED DATA OR CENSOR BY INDIVIDUAL THRESHOLD}

\textbf{DET HER ER NOGET VÆRRE ROD. DER ER FORSVUNDET NOGET LASSO TEKST. GÅ TILBAGE OG FIND DET IGEN}
Under the deviance optimization criterion the median number of non-zero
coefficients was 33 (range 18 to 52). In order to calculate the sequential
rank agreement for the Lasso we therefore applied our proposed Algorithm \ref{sra-algorithm}
with $B = 1000$ and $P=5000$.

Thus, in our application the median number of censored predictors
across the 1000 sub-samples was 5 (range 1 to 11). Even though the
number of censored predictors is very small compared to the total
number of predictors, the effect on the sequential rank agreement is
substantial and with the artificial censoring the shape of the curves
is as expected, starting low and then increasing. Also, the effect of
the censoring caused by the sparsity of the Lasso is clearly visible
in the agreement curves.
\textbf{ANDREAS: man kan desvaerre ikke se hvor Lasso agreement forsvinder
  (ved list-depth 5?) og om den er laver for list depth 1-5 end rigde
  eller om den er kun laver for list depth 1-2. maaske kan du
  beskrive dette i teksten?}

@@

Most curves start low and then increase indicating higher agreement
among high risk patients. This is expected if we rank the individuals
according to highest risk of disease. However, it is also expected
that individuals with very low risk also show high agreement. In this
case we order the patients according to (high) risk prediction but we
could essentially also have reversed the order to identify the
patients that have low risk prediction.  An exception is the risk
prediction agreement for the Lasso tuned with the AUC criterion which
shows very low agreement among the high values of the predicted
risks. The reason is that optimizing the penalty parameter with
respect to the AUC criterion tends to favor a very high penalty value
causing only a single predictor to be selected in each of the 1000
iterations. This results in a lack of generalizability to the DACOVA
data which gives rise to the higher disagreement in the predicted risks. 
In the extreme case where the penalty becomes so high
that none of the predictors are selected by the Lasso, the sequential
rank agreement for the predicted probabilities becomes undefined since
all the ranks will be ties.

@@

Lasso shows similar agreement for ranking the risk predictions than
PLS-DA (unless with the AUC criterion), and poor agreement for ranking
predictors. This reason for the latter is the high auto-correlation
between the intensities in the mass spectra which leads to
collinearity issues in the regression models. It is well-known that
variable selection with the Lasso penalty does not preform very well
when the predictors are highly correlated. The collinearity does,
however, not affect the agreement of the risk predictions, as seen on
the left panel of Figure \ref{fig:app1}, since it is not so important
which specific variable is selected from a group of highly correlated
predictors when the purpose is risk predictions.

It appears that Ridge regression tuned with the AUC criterion achieves
the best performance with respect to the stability of ranking the
individual predicted risk probabilities. It must, however, be stressed
that the sequential rank agreement in this application is only
concerned with the agreement of the risk predictions across
sub-samples and not with the actual accuracy of the risk
predictions. Thus, we also computed the AUC and Brier scores for the different models
based on the DACOVA data. The distributions across the $1000$
sub-samples for a selection of the models is shown in the right panel
of Figure \ref{fig:app2}.  Here we see that PLS-DA attains the highest
AUC values with a median value of $0.71$ while the Ridge model with
the AUC criterion attains a median AUC of $0.49$. This implies that
while Ridge regression optimized with respect to the AUC criterion
achieves the best sequential rank agreement it performs similar to a
random coin toss with respect to classifying the DACOVA patients.
\textbf{Comment on Brier scores}
Both concerns are of course important in practice and for this data set 
it appears thatPLS-DA is the overall best method taking both agreement 
and classification accuracy into account.

@@

Our applications show that XXX
Through simulations we showed that XXX --- afventer Andreas.

>>>>

Sequential rank agreement can be generalized to incomplete/censored
lists in the following way. Let $\Lambda_P$ be a set of $P$ items and
$\Lambda_l\subset \Lambda_P$ the subset of $d_l$ items that occur in
list $l$.  The case where all lists are censored at depth $d$
corresponds to $d_1=\cdots=d_L=d$. For incomplete/censored lists the
rank function becomes
\begin{equation}
  \widetilde R_l(X_p) = \left\{\begin{array}{cl} \{R_l(X_p)\} & \text{ for } X_p\in \Lambda_l \\
      \{d_l+1,\dots,P\} & \text{ for } X_p \not\in \Lambda_l\end{array}\right.
\end{equation}
where we only know that the rank for the unobserved items in list $l$
must be larger than the largest rank observed in that list.

@@

Alternatively, we can remove the restriction on the independence among
the methods by only requiring that there is no information in the
ranking but that the rankings are all based on applying the
method/approaches to the same data
\begin{eqnarray*}
\widetilde H_0 & :&  \text{The list rankings are based on data containing}\\
& &   \text{no association to the outcome.}
\end{eqnarray*}
$H_0$ is a quite unrealistic null hypothesis but we can easily obtain
realizations from that null hypothesis simply by permuting the items
within each list and then computing the sequential rank agreement for
the permuted lists. In the uncensored case each experiment contains
$L$ lists of random permutations of the numbers $1,\dots,P$. For the
censored/incomplete case we first permute the numbers $1,\dots,P$ and
then censor list $l$ at $d_l$. The sequential rank agreement curve
from the the original lists can then be compared to the pointwise rank
agreements obtained under $H_0$.

To obtain the distribution under $\widetilde H_0$ the idea is to repeat
many times the ranking procedures in uncorrelated data.  Thus, we
first permute the outcome variable of the data set on which the
rankings are based. This removes any association between the predictor
variables and the outcome. Then we apply the methods to the permuted
data set to generate $L$ new rankings and subsequently compute the
sequential rank agreement. Note that we only permute the outcomes and
preserve the structure of the candidate predictors. The randomization
approach requires that we have the original data available and as such
it may not be possible to evaluate $\widetilde H_0$ in all situations.

If the sequential rank agreement for the original data lies
substantially below the distribution of the sequential rank agreements
obtained under either $H_0$ or $\widetilde H_0$ then this suggests that
the original ranked lists agree \emph{more} than what we would expect
in data with no information.

For the analysis of the Golub data described in Section
\ref{sec:amfol}, Figure~\ref{fig:example1} shows the empirical
distributions of sequential rank agreement under $H_0$ and
$\wildetilde H_0$ each based on $400$ permutations. Not surprisingly, the
sequential rank agreement under $\widetilde H_0$ is lower than the
sequential rank agreement under $H_0$ because the four methods used to
rank the data ($t$ test, logistic regression, elastic net, and MIC)
generally tend to identify (and rank high) similar predictors even if
there are only spurious associations. The two bottom panels in
Figure~\ref{fig:example1} also indicate that the observed sequential
rank agreement (the black line) is better than what would be expected
by chance for data that contain no information since it lies below the
reference areas. However, the censored data also suggests that there
may be at most 1 or 2 ranked items towards the top of the lists that
yield a result better than what would be expected (the bottom-right
plot).

It is important to stress that neither $H_0$ nor $\widetilde H_0$ are
related to questions regarding the association between the outcome and
the predictors in the data set. Both hypotheses are purely considering
how the rankings agree in a situation where there is no relevant
information available in the data used for creating the rankings.

@@

A not expected shape of the agreement curves is seen for the Ridge
models for all three tuning criteria. They all show higher
disagreement in the top of the lists followed by a decrease. The
reason behind this behavior is rather subtle. Looking at the
distribution of the absolute value of the regression coefficients we see that a large
proportion of them are numerically very close to zero and have almost
equal absolute value. This is a general feature of the Ridge models in
this data set and seen for all the 1000 trained models.
This implies that when predictors are ranked according to the
magnitude of their coefficients, their actual order becomes more
uncertain and more close to a random permutation. This problem can be
alleviated by truncating all predictors with absolute coefficient values 
below a given threshold thereby introducing an artificial censoring of the lists. 
For the Ridge models tuned with the deviance criterion, Figure \ref{fig:app2} (left) shows the
the sequential rank agreement where for each of the 1000 trained
models the predictors were artificially censored when their absolute
coefficient value was lower than the $0.1\%$ quantile of the 5000
absolute coefficient values. The curve was calculated
using Algorithm \ref{sra-algorithm} with $B=1000$ and $P=5000$. The corresponding 
curve from Figure \ref{fig:app1} (right panel) is shown for comparison.
Even though the number of censored predictors is very small compared to the total
number of predictors, the effect on the sequential rank agreement is
substantial and with the artificial censoring the shape of the curves
is as expected, starting low and then increasing.

Looking at the agreement curves for the Lasso models in Figure \ref{fig:app1} 
(right) we clearly see the effect of the sparsity inducing penalization giving 
rise to censored lists. These curves were similarly calculated using \ref{sra-algorithm} 
and $1000$ random permutations. Under the deviance optimization criterion the 
median number of non-zero coefficients was 33 (range 16 to 50) and for the 
class accuracy criterion 14 (range 4 to 56). These values correspond to
the list depths where the agreement curves become constant as a result of
the subsequent censoring.

@@

Most curves start low and then increase indicating higher agreement
among high risk patients. This is expected if we rank the individuals
according to highest risk of disease. However, it is also expected
that individuals with very low risk also show high agreement. In this
case we order the patients according to (high) risk prediction but we
could essentially also have reversed the order to identify the
patients that have low risk prediction.

An exception is the risk prediction agreement for the Lasso tuned with the AUC criterion which
shows very low agreement among the high values of the predicted
risks. The reason is that optimizing the penalty parameter with
respect to the AUC criterion tends to favor a very high penalty value
causing only a single predictor to be selected in each of the 1000
iterations. This results in a lack of generalizability to the DACOVA
data which gives rise to the higher disagreement in the predicted risks. 
In the extreme case where the penalty becomes so high
that none of the predictors are selected by the Lasso, the sequential
rank agreement for the predicted probabilities becomes undefined since
all the ranks will be ties.

@@

Lasso shows similar agreement for ranking the risk predictions than
PLS-DA (unless with the AUC criterion), and poorer agreement for ranking
predictors. This reason for the latter is the high auto-correlation
between the intensities in the mass spectra which leads to
collinearity issues in the regression models. It is well-known that
variable selection with the Lasso penalty does not preform very well
when the predictors are highly correlated. The collinearity does,
however, not affect the agreement of the risk predictions, as seen on
the left panel of Figure \ref{fig:app1}, since it is not so important
which specific variable that gets selected from a group of highly correlated
predictors when the purpose is risk predictions.

It appears that Ridge regression tuned with the AUC criterion achieves
the best performance with respect to the stability of ranking the
individual predicted risk probabilities. It must, however, be stressed
that the sequential rank agreement in this application is only
concerned with the agreement of the risk predictions across
sub-samples and not with the actual accuracy of the risk
predictions. Thus, we also computed the AUC values for the different models
based on the DACOVA data. The distributions across the $1000$
sub-samples for a selection of the models is shown in the right panel
of Figure \ref{fig:app2}.  Here we see that PLS-DA attains the highest
AUC values with a median value of $0.70$ while the Ridge model with
the AUC criterion attains a median AUC of $0.49$. This implies that
while Ridge regression optimized with respect to the AUC criterion
achieves the best sequential rank agreement, it performs similar to a
random coin toss with respect to classifying the DACOVA patients.
In practice both concerns are of importance.

@@

\textbf{Our applications show that XXX
Through simulations we showed that XXX --- afventer Andreas.}
